Evaluating Russia's Preparedness and Response to the COVID-19 Pandemic: An Analysis of Excess Death Rates and the Global Health Security Index

### 2024-02-14 paper/paper.qmd
proof-reading changes (until discussions)

---
title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations. This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates. Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries. This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The COVID-19 pandemic has led to an enormous loss of life over the world and has presented an unprecedented challenge to public health, foreign policy and has additionally had an adverse effect on countries' economies around the world, over the past 5 years [@citeWHOeffect]. There have been over 774 million reported confirmed cases of the virus being contract and the outbreak has caused over 7 million deaths, which are claimed to be "under counted" and "under reported", suggesting that the actual number is even higher around 18.2 - 33.5 million based on the excess mortality estimates [@citeWikiCOVID]. This excessively high number of deaths has alerted the world to assess how countries assumed to be well-prepared to tackle such outbreaks were affected to this extent.

There were a subset of countries, namely India, the United States of America, Brazil, Russia and Mexico which were impacted the most in terms of having the highest cumulative total deaths [@citeThinkGlobalHealth]. According to the global data, Russia's COVID-19 related deaths since the start of the pandemic have been over 230,000. Russia also holds the highest mortality rate in Europe and the second-highest rate in Asia [@citeCarnegieRussia]. For these reasons and more, Russia and its residents have suffered a lot over the past few years, wherein the adverse impact of the outbreak has lead to a plethora of impacts on the country's economy and standard of living.

This paper will follow a reproduction of Jennifer B. Nuzzo and Jorge R. Ledesma's paper [@citeRepPaper], where they used The Global Health Security (GHS) Index an indicator of how well prepared or "secure" a country is against epidemics or pandemics like the COVID-19 outbreak. We in-turn will use these findings and apply a Russian-facing lens to discuss the impact of the COVID-19 outbreak on Russia and what could have been done better and can be improved for the future in order for the impact of any upcoming pandemics can be lessened. Our paper aims to replicate their claim of how well-prepared countries did so poorly during the COVID-19 outbreak, where the Nuzzo and Ledesma have concentrated solely on the United States but in this paper we look at Russia with the same regards and analyse the impact of the pandemic on the death toll, and finally provide for ways to improve the laws, regulations and plans to tackle such outbreaks in the future. Our reproduction found out that how ever well prepared Russia seemed to be, it experienced an unprecedented rise in the death rate compared to other countries, and its excess death rate was one of the highest amongst all countries. This leads us to make an important judgement that better planning and regulations can help reduce the impact of such outbreaks in the future, which will be studied in this paper.

This paper is structured using the following sections: Data, Results and Discussion. In the Data (@sec-data) section, the data source of the datasets from the paper being replicated [@citeRepPaper] is discussed and the measurement and data cleaning process is outlined. In the Results (@sec-results) section, the paper summarizes the data findings and presents relevant maps and plots in order to study the trends. The paper ends with the Discussion (@sec-discussion) section, where the findings of the paper have been analysed and delved deeper into by studying the measures that can be taken to prevent the country from experiencing such an extensive effect of future pandemics. Additionally, the limitations and a further scope for the paper has been discussed here and concluded in the end.

@@

As mentioned above, the data analysed in this paper is from the reproduction package of the paper by Nuzzo and Ledesma [@citeRepPaper]. The data was cleaned and analysed using the open source `R` programming language [@citeR]. R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf]. Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper. The estimands for this analysis are preparedness of countries to tackle pandemic, specifically Russia, and their degree of success in controlling the pandemic. To estimate these, we utilize metrics such as the Global Health Security (GHS) Index and the excess death rate respectively.

In the following sections, we will discuss the raw data in (@sec-data-raw), the data cleaning process in (@sec-data-clean) and then move on to discussing the measurement in (@sec-data-measurement).

@@

This paper will replicate the data that was obtained from the replication package of the paper [@citeRepPaper]. Specifically, in the paper we use 3 datasets that will help us analyse the claim of Russia not being able to control the wave of the COVID-19 pandemic well. In general, there is a lot of data available on the effect of COVID-19 on countries, however, all the data we needed for this paper was included in the paper by Nuzzo and Ledesma.

The first raw dataset includes data on the Global Health Security index for all countries. This dataset has a plethora of data on indices that are not relevant to us for this paper. Additionally, the dataset contains data for years 2019 and 2021. For our analysis, we only require data for the GHS index of countries for 2021.

The second raw dataset includes data on all cause death rates for all countries from 1960-2021. Within this dataset, a plethora of information exists, the majority of which proved extraneous for the purposes of our paper. Specifically, data preceding the year 2010 is deemed unnecessary for our analysis, since we are only focusing on COVID-19 we will use this dataset to study the trend in death rates for countries around the world from 2010-2021.

The third raw dataset originates from the Institute of Health Metrics and Evaluation (IHME) and provides information on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for all countries worldwide. Similar to the previous datasets, this dataset also contains extraneous information that is irrelevant to our study. However, we addressed this issue through the data cleaning process outlined in Section @sec-data-clean.

In the next section (@sec-data-clean), we will outline the data-cleaning process and also show the structure of the cleaned data.

@@

As stated above in @sec-data-raw, we start with the first dataset on GHS indices for all countries in the world. @tbl-cleaned-ghs shows the first 6 countries and their GHS indices. Here we only have 2 variables, namely the Country Name and the respective GHS index. We cleaned this dataset by selecting only the two columns for the respective variables and filtering the data for the year of 2021.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-ghs
#| tbl-cap: "Cleaned Data showing GHS indices for first 6 countries"
#| tbl-cap-location: bottom
#| warning: false

@@

Second, we have a raw dataset on all cause death rates for all countries from 1960-2021. We cleaned this dataset by selecting a subset of countries that helped us analyse how Russia fared against other countries in death ratesover the years. @tbl-cleaned-death-rates shows the structure of the cleaned data after we filtered it for the time period of 2010-2021 for India, South Korea, Russia and the United States.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-death-rates
#| tbl-cap: "Cleaned Data showing all cause death rates for 4 countries"
#| tbl-cap-location: bottom
#| warning: false

@@

The final raw dataset, as mentioned in @sec-data-raw, that we have is the IHME data on the COVID death rate, excess deaths, excess death rate and ratio of excess death rate over COVID death rate for all countries in the world. In this paper we only needed the two variables excess death rate and the country name, however for better analysis of Russia's ability to control the pandemic's effect, we combined this data with the respective GHS index as shown in @tbl-cleaned-combined. Hence, there are three variables, the country name, and the respective excess death rate and GHS index.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-combined
#| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for first 6 chosen countries"
#| tbl-cap-location: bottom
#| warning: false

@@

Now that we have cleaned our raw datasets and have generated clean tables for analysis, we will go on to the Results Section (@sec-results) and delve deeper into studying the effect of the COVID-19 pandemic on Russia and if it was impacted significantly more than other countries.

@@

@@

The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's death rates trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).

@@

As explained in @sec-data-raw, the GHS index is an indicator of how well prepared a country is to tackle epidemics or pandemics like the COVID-19 outbreak. The GHS index gives one an idea of how the sound health care system  as well as the rules and regulations of a certain country are in the context of facing pandemics. Generally, the higher the GHS index of a country, the more well-prepared and secure they are. @fig-map shows a representation of the world map with countries shaded on the basis of their GHS indices. This map is a replication of figure 1 in Nuzzo and Ledesma paper [@citeRepPaper] that is being replicated in this report.

@@

We can observe here that there are countries like the United States and Australia whose GHS indices are one of the highest, implying that they were deemed extremely well prepared for epidemics. Then we see countries like Russia and China who were, although moderately when compared to the aforementioned countries, but still well prepared to tackle pandemics. Finally, we see countries such as Venezuela and Yemen which were said to be the least secure of all.

@fig-map provides this report with a background on how secure Russia was in tackling epidemics and pandemics, which as seen in the map, it was in a good position to face such outbreaks where numerous countries were deemed less prepared than Russia. This will aid us in drawing meaningful analyses throughout the paper, specifically in @sec-results-death-rates, sec-results-death-rates and @sec-discussion.

@@

Now we move on to analyzing the trends in the all cause death rates per 1000 population over 2010-2021, which includes COVID deaths as well as general deaths which were not caused by the outbreak, and the trend in the excess death rates per 100 population in 2020 and 2021. The excess death rates in this section were calculated by finding a difference between the expected death rate and the actual death rate for that year.

@fig-all-cause-death-rate shows us the trend for the all cause death rates per 1000 population over 2010-2021 for India, South Korea, Russia and the United States using the data presented in @tbl-cleaned-death-rates. We observe here that from the years 2010-2019, there is a general gradually decreasing trend in the death rates in the case of Russia and India, and a gradually increasing trend in the death rates in the case of South Korea and the United States. The dashed line represents the line of best fit for the trend from 2010-2019.

@@

We see here in @fig-all-cause-death-rate that there is a sudden increase in the death rate in 2020 and 2021, and the increase in Russia's death rate is much higher than that of the other countries. The essential point to note here is that in the case of India, which has a GHS index lower than that of Russia, the rising trend observed in 2020 and 2021 was still not as excessive as in the case of Russia. The dotted line represents the line of best fit for the trend from 2010-2021. As we go from comparing Russia to South Korea, we see that in the case of South Korea, both the dashed (2010-2019) and the dotted (2010-2021) are very close together implying that the country did not experience an unanticipated number of deaths during COVID. However, in the case of Russia, we see the two lines diverging implying that there were unforeseen deaths during that period.

## Excess Death Rates {#sec-results-excess-rates}
This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021. As mentioned above, the excess death rates were calculated by extrapolating the dashed line of best-fit from years 2010-2019 in @fig-all-cause-death-rate representing the trend in all cause death rates to 2020 and 2021. This way we were able to find the expected death rate for each of the four countries and subsequently we were able to plot the difference between the expected death rate and the actual death rate called the excess death rate for 2020 and 2021 for all four countries in @fig-excess-death-rate.

@@

We see in @fig-excess-death-rate the massive difference between the excess death rates for Russia and the other countries. Comparing Russia to South Korea we see that, as seen in @fig-all-cause-death-rate, South Korea was very faintly affected by the COVID-19 pandemic, however Russia saw a sudden rise in deaths from 2019 to 2021. This shows us that the observations here are analogous to the plot in @fig-all-cause-death-rate. When we compare Russia with India, which as aforementioned was deemed a less secure country in the case of epidemics, we see that the excess death rates for Russia in both 2020 and 2021 are much higher than those for India. This observation will further aid us in making meaningful discussions in @sec-discussion.

@@

Finally, we will now explore the relationship between the excess death rate seen above and the GHS index for a subset of countries per 100,000 population. @fig-excess-death-rate-ghs shows us a subset of countries scattered around the plot according to their excess death rate and the GHS index shown in the Appendix @tbl-appendix. The green line shows the general decreasing trend which implies that as the GHS index increases, the excess death rate tends to decrease, i.e. the excess death rate for more "secure" or well-prepared countries is generally lower.

Russia, labelled in red, is seen to be significantly higher than the line of best fit, which implies that Russia's excess death rate was much worse than it should have been. We observe that countries like Brazil and Indonesia, which have approximately the same GHS index as Russia, are nearer to the line of best fit than Russia. India, although having a lower GHS index than Russia is seen to have a much lower than general excess death rate.

@@

The implications from this figure will be further studied in @sec-discussion.

>>>>

---
title: "Evaluating Russia's Preparedness and Response to the COVID-19 Pandemic: An Analysis of Excess Death Rates and the Global Health Security Index"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "This study investigates the unexpected surge in death rates during the COVID-19 pandemic in Russia, a country considered well-prepared according to the Global Health Security (GHS) Index. Building upon Nuzzo and Ledesma's work, we analyze the impact of COVID-19 in Russia, incorporating factors such as the GHS Index and excess death rates. Our findings reveal a substantial increase in excess death rates in Russia, diverging from the lower rates observed in nations deemed well-prepared. This research contributes to our understanding of the repercussions of the COVID-19 outbreak in Russia, highlighting areas for improvement in prepared countries to mitigate the effects of future pandemics."
format: 
  pdf:
    tbl-pos: "h"
    fig-pos: "h"
toc: true
number-sections: true
bibliography: references.bib
---

@@

The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years (World Health Organization [@citeWHOeffect]). With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates (WikiCOVID [@citeWikiCOVID]). The staggering number of deaths has prompted a critical examination of countries presumed to be well-prepared to handle such outbreaks and the factors contributing to their extensive impact.

Among the most severely affected nations, including India, the United States of America, Brazil and Mexico, Russia, a nation deemed well-prepared according to various metrics, was specifically chosen for this study due to its unexpected and disproportionately high death toll in the face of the pandemic. With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia (ThinkGlobalHealth [@citeThinkGlobalHealth]; Carnegie Russia [@citeCarnegieRussia]). This selection aims to uncover unique insights into the challenges faced by ostensibly prepared nations.

This paper replicates the methodology employed by Jennifer B. Nuzzo and Jorge R. Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 (RepPaper [@citeRepPaper]). The choice of GHS is deliberate, considering its comprehensive evaluation of various health security capacities. This metric offers a nuanced understanding of a country's ability to prevent, detect, and respond to public health emergencies, making it a suitable choice over other metrics. However, our focus extends to Russia, providing insights into the impact of the outbreak and proposing improvements for future pandemic mitigation efforts. Despite Russia's apparent preparedness, our reproduction reveals an unprecedented rise in the death rate compared to other countries, with one of the highest excess death rates globally. This prompts a crucial assessment that better planning and regulations are imperative to mitigate the impact of such outbreaks, a key aspect explored in this paper.

The structure of this paper comprises three sections: Data, Results, and Discussion. In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper (RepPaper [@citeRepPaper]). The Results section (@sec-results) summarizes the findings, presenting maps and plots to illustrate trends. The Discussion section (@sec-discussion) delves into the analysis of our findings, exploring measures to prevent extensive effects of future pandemics on the country. Additionally, we discuss limitations and propose avenues for further research, concluding the paper.

@@

The dataset under examination in this study originates from the replication package associated with Nuzzo and Ledesma's paper [@citeRepPaper]. Employing the open-source `R` programming language [@citeR], we conducted the cleaning and analysis procedures, leveraging several R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth], and `sf` [@citeSf].

Our data cleaning and analysis processes replicate those of the original paper, ensuring consistency in the datasets used. In terms of estimand, the primary focus of this analysis is to assess the preparedness of countries, specifically Russia, in addressing pandemics, and to gauge their success in controlling the associated impact. Key estimators include the Global Health Security (GHS) Index, offering insights into overall preparedness, and the excess death rate, serving as an indicator of the pandemic's impact.

The subsequent sections will provide a comprehensive overview of the raw data in (@sec-data-raw), delve into the intricacies of the data cleaning process in (@sec-data-clean), and subsequently, explore the measurement aspects in (@sec-data-measurement).

@@

This study replicates the data sourced from the replication package of Nuzzo and Ledesma's paper [@citeRepPaper]. Specifically, we utilize three datasets to investigate the assertion that Russia struggled to effectively manage the surge of the COVID-19 pandemic. Despite the wealth of data available on the global impact of COVID-19, the comprehensive datasets provided by Nuzzo and Ledesma serve as the primary source for our analysis.

The first raw dataset encompasses the Global Health Security index for all countries, offering a wide array of indices, some of which are not pertinent to our study. Notably, this dataset spans the years 2019 and 2021; however, our analysis focuses exclusively on the GHS index for the year 2021.

The second raw dataset provides data on all-cause death rates for all countries from 1960 to 2021. The "all-cause death rate" refers to the overall mortality rate in a given population, considering deaths from all possible causes. It is a comprehensive measure that includes deaths from various diseases, accidents, and other factors. While this dataset contains extensive information, we restrict our analysis to data from 2010 onwards, as earlier years are deemed unnecessary for our COVID-19-focused investigation.

The third raw dataset, sourced from the Institute of Health Metrics and Evaluation (IHME), provides us information on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for countries globally. Similar to the preceding datasets, unnecessary information was present, requiring thorough data cleaning, a process detailed in Section @sec-data-clean.

The subsequent section (@sec-data-clean) will show the data cleaning process, elucidating the structure of the refined dataset employed in our analysis.

@@

As outlined in @sec-data-raw, our initial step involves the first dataset concerning Global Health Security (GHS) indices for all countries globally. Presented in @tbl-cleaned-ghs are the first six countries along with their corresponding GHS indices. The dataset has been refined to include only two variables, namely Country Name and the respective GHS index. Cleaning involved the selection of these two columns and filtering the data specifically for the year 2021.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-ghs
#| tbl-cap: "Cleaned Data showing GHS indices for first 6 countries"
#| warning: false

@@

Moving to the second dataset covering all-cause death rates for countries from 1960 to 2021, the cleaned version is showcased in @tbl-cleaned-death-rates. To analyze Russia's position relative to other nations in terms of death rates, we filtered this dataset for the years 2010-2021, focusing on countries like India, South Korea, Russia, and the United States. The refined dataset structure is highlighted, enabling a clearer examination of death rate trends.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-death-rates
#| tbl-cap: "Cleaned Data showing all cause death rates for 4 countries"
#| warning: false

@@

The last raw dataset, detailed in @sec-data-raw, originates from the Institute of Health Metrics and Evaluation (IHME), providing data on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for countries worldwide. While our primary interest was in the excess death rate and the country name, for a more comprehensive analysis of Russia's pandemic management, we integrated this data with the respective GHS index, showcased in @tbl-cleaned-combined. Consequently, this dataset comprises three variables: the country name, the respective excess death rate, and the GHS index.

```{r}
#| echo: false
#| eval: true
#| label: tbl-cleaned-combined
#| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for first 6 chosen countries"
#| warning: false

@@

With the completion of the data cleaning phase and the generation of refined tables for analysis, our next section will delve into the Measurement (@sec-data-measurement), offering an in-depth exploration of how the phenomena observed translated into entries in our dataset. Understanding this process is crucial for interpreting patterns and drawing meaningful conclusions from the data analysis results.

@@

The subsequent section (@sec-results) will present the results of our analysis, offering insights into Russia's preparedness and response to the COVID-19 pandemic, and the implications of the findings.

@@

The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's all-causes mortality rate trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).

@@

As outlined in @sec-data-raw, the Global Health Security (GHS) index serves as a metric indicating a country's preparedness to address epidemics or pandemics, such as the COVID-19 outbreak. Reflecting both the robustness of the healthcare system and the regulatory framework, a higher GHS index suggests a more secure and well-prepared nation. @fig-map presents a world map illustrating countries shaded based on their GHS indices. This replication of figure 1 from Nuzzo and Ledesma's paper [@citeRepPaper] provides an overview of global preparedness.

@@

Noteworthy observations reveal countries like the United States and Australia boasting some of the highest GHS indices, indicating exceptional preparedness for epidemics. Countries like Russia and China, while moderately prepared compared to the former, still exhibit a substantial level of readiness. Conversely, nations such as Venezuela and Yemen rank among the least secure.

@fig-map provides us essential contextual information on Russia's preparedness in the face of epidemics and pandemics. Positioned favorably, Russia emerges as well-prepared compared to numerous countries deemed less secure. This background sets the stage for meaningful analyses throughout the paper, particularly in @sec-results-death-rates and @sec-discussion.

@@

We proceed to examine the trends in all-cause death rates per 1000 population from 2010 to 2021, encompassing both COVID-related deaths and non-COVID-related fatalities, as well as the trend in excess death rates per 100 population in 2020 and 2021 in the subsequent section. @fig-all-cause-death-rate presents the trends in all-cause death rates per 1000 population from 2010 to 2021 for India, South Korea, Russia, and the United States, utilizing data from @tbl-cleaned-death-rates. From 2010 to 2019, a general gradual decrease in death rates is evident for Russia and India, juxtaposed with a gradual increase observed in South Korea and the United States. The dashed line represents the line of best fit for the years 2010-2019.

@@

From @fig-all-cause-death-rate, a notable surge in death rates is observed in 2020 and 2021, particularly pronounced in Russia compared to other countries. It is noteworthy that despite India's lower GHS index compared to Russia, the observed increase in death rates during 2020 and 2021 was not as substantial as in Russia. The dotted line represents the trend line for the years 2010-2021. Comparing Russia to South Korea, the dashed (2010-2019) and dotted (2010-2021) lines closely align for South Korea, indicating a relatively consistent trend without a significant increase in unexpected deaths during COVID-19. Conversely, for Russia, the divergence between the two lines implies a notable increase in unforeseen deaths during this period.

## Excess Death Rates {#sec-results-excess-rates}
Considering the likelihood of undercounting COVID-19 deaths, our focus shifted to analyzing excess death rates for the years 2020 and 2021 to provide a comprehensive measure of the pandemic's overall mortality impact. Drawing inspiration from Nuzzo and Ledesma's methodology (@citeRepPaper), which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model. Extrapolating the dashed line of best-fit from 2010-2019 to 2020 and 2021 allowed us to determine the expected death rate for each country. The subsequent plotting of the difference between expected and actual death rates, termed the excess death rate, unveiled insights into the pandemic's impact on mortality.

Our calculated values of excess death rates are presented in the figure labeled "Excess Death Rate" (@fig-excess-death-rate), showcasing the rates for Russia, India, South Korea, and the United States in 2020 and 2021. Simultaneously, the figure labeled "All-Cause Death Rate" (@fig-all-cause-death-rate) illustrates the trend in all-cause death rates for the same period. This simple models provides a nuanced understanding of the excess death rates, contributing valuable insights into the broader mortality implications of the COVID-19 pandemic.

@@

In @fig-excess-death-rate, a considerable contrast is evident, particularly highlighting the substantial difference in excess death rates for Russia compared to other countries. The comparison with South Korea, as depicted in @fig-all-cause-death-rate, reinforces the minimal impact of the COVID-19 pandemic on South Korea, while Russia experienced a notable surge in deaths from 2019 to 2021. Comparing Russia to India, designated as less secure in the context of epidemics, the excess death rates for Russia in both 2020 and 2021 significantly exceed those for India. This observation will serve as a crucial point of reference in our ensuing discussions in @sec-discussion.

@@

Finally, we delve into the relationship between the excess death rate and the Global Health Security (GHS) index for a subset of countries per 100,000 population. @fig-excess-death-rate-ghs visually represents this subset of countries scattered on the plot based on their excess death rate and GHS index, as detailed in Appendix @tbl-appendix. The green line signifies a general decreasing trend, suggesting that as the GHS index increases, the excess death rate tends to decrease – indicating that more "secure" or well-prepared countries generally experience lower excess death rates.

Russia, highlighted in red, stands notably above the line of best fit, indicating that its excess death rate exceeded expectations. In contrast, countries like Brazil and Indonesia, with comparable GHS indices to Russia, align more closely with the line of best fit. India, despite having a lower GHS index than Russia, exhibits a substantially lower excess death rate.

@@

@fig-excess-death-rate-ghs prompts further exploration in @sec-discussion, where we will delve into the implications and potential factors contributing to Russia's disproportionately higher excess death rate compared to countries with similar or even lower GHS indices.
